\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-01-15 paper.tex
update markers

But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule as is for risk prediction, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow in a context where we want to perform real-time risk prediction. 
Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction -- a natural and appropriate marker rule of the lights model for subject $i$ being in subgroup $k$ would rather be 
\begin{equation}
  \label{eq:marker-def}
  \hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)},
\end{equation}
with 
\[ \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
where $\hat S$ is obtained with Algorithm~\ref{alg:construct-set-S} using $\hat D$ (that is $S^{(w_{max})}$ for the last iteration $w_{max}$), and the density $f$ in the rigth hand side of the previous equation being defined in~\eqref{eq:ind-hyp}.
Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:hat_pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability 
\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i, x_i].\]

@@

We detail in this section the metric considered to evaluate risk prediction performances. Let us recall that in practice, one considers the case $K=2$, see Section~\ref{sec:K=2}. 
In this context, the predictive marker for the lights model is
\[\hat \cR_i^\text{lights} = \dfrac{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i) \big) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}.\]
Then, let us denote by $\hat \cR$ the marker under study in the general case, and assume that it is measured once at $t = t^{max}$.
A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
\begin{equation*}
  \cC =\P[\hat \cR_i > \hat \cR_j | T^\star_i < T^\star_j],
\end{equation*}
with $i \neq j$ two independent subjects (which does not depend on $i, j$ under the i.i.d. sample hypothesis). 
In our case,  $T^\star$ is subject to right censoring, so one would typically consider the modified $\cC_\tau$ defined by
\begin{equation*}
  \cC_\tau =\P[\hat \cR_i > \hat \cR_j | T_i < T_j , T_i < \tau],
\end{equation*}
with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}. A Kaplan-Meier estimator for the censoring distribution leads to a nonparametric and consistent estimator of $\cC_\tau$ \citep{uno2011c}, already implemented in the \texttt{Python} package \texttt{lifelines}.
Hence in the following, we consider the C-index metric to assess performances.

@@

The first model we consider as a baseline is the well known Cox PH model with time-independent features, also known in the joint modeling context as the ``landmark'' model, in which we include basic time-independent features extracted from longitudinal processes and computed at the prediction time [REF].
In the Cox PH model introduced in \citet{Cox1972JRSS}, a parameter vector $\beta$ is estimated by minimizing the partial log-likelihood given by
\begin{equation*}
  \ell_n^{\text{cox}}(\beta) = n^{-1} \sum_{i=1}^n \delta_i \big( x_i^\top \beta - \log \sum_{i' : t_{i'} \geq t_i} \text{exp}(x_{i'}^\top \beta) \big).
\end{equation*}
For each time-dependant feature, we include subject-specific random effects, values of longitudinal processes at baseline and at time $t^{max}$, slope of longitudinal processes at time $t^{max}$ and area under of longitudinal processes from baseline up to $t^{max}$. The predictive marker for a new subject $i$ is the regular risk score~\citep{therneau2000cox}, that is $\hat \cR_i^{\text{cox}} = \exp (x_i^\top\hat\beta^{\text{cox}})$, with 
\[\hat\beta^{\text{cox}} \in \argmin_{\beta \in \R^\vartheta} -\ell_n^{\text{cox}}(\beta).\]
We use the \texttt{Python} package \texttt{tick} for the estimation of the previous quantity.
Ties are handled via the Breslow approximation of the partial likelihood \citep{breslow1972contribution}.

@@

We then consider a penalized version of the Cox model, including now much more time-independent features automatically extracted from longitudinal processes using the \texttt{Python} package \texttt{tsfresh}.
The predictive marker for the a subject $i$ is $\hat \cR_i^{\text{cox EN}} = \exp (x_i^\top\hat\beta^{\text{cox EN}})$, with 
\[\hat\beta^{\text{cox EN}} \in \argmin_{\beta \in \R^\vartheta} -\ell_n^{\text{cox}}(\beta) + \xi \big( (1-\eta)\norm{\beta}_1 + \frac{\eta}{2} \norm{\beta}_2^2 \big)\] 
the estimate obtained using elastic net penalization (still with the \texttt{tick} package), and
where $\xi$ is chosen by the a 10-fold cross-validation procedure, for a given $\eta \in [0, 1]$.

@@

A classical extension of the Cox model supposes that features depend on time~\citep{sueyoshi1992semiparametric} (and then loosing the risk proportionality property), that is with our notations
\[\lambda_i(t) = \lambda_0(t) \exp\big(y_i(t)^\top \beta \big). \]
Using the \texttt{survival} \texttt{R} package~\citep{zhang2018time}, blabla description...

@@

We consider a multivariate version of JLCMs implemented in the \texttt{R} package \texttt{lcmm}~\citep{2017_lcmm}.
Contrary to the lights model, there is no shared associations between the longitudinal models and the survival model. Given the group membership, each submodel is assumed to be independent. In this context, the predictive marker is
\[\hat \cR_i^\text{lcmm} = \dfrac{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i)\big) \hat f(t^{max}_i, y_i | b_i, G_i=0 ; \hat \theta)}.\]

@@

In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. For a new subject $i$, the predictive marker associated with this model is naturally given by the corresponding risk score
\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, \hat \beta_k^l, b_i^l) \Big\}.\]

>>>>

But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule as is for risk prediction, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow in a context where we want to perform real-time risk prediction. 
Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction -- a natural and appropriate marker rule of the lights model for subject $i$ being in subgroup $k$ would rather be 
\begin{equation}
  \label{eq:marker-def}
  \hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)},
\end{equation}
with 
\[ \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
where $\hat S$ is obtained with Algorithm~\ref{alg:construct-set-S} using $\hat D$ (that is $S^{(w_{max})}$ for the last iteration $w_{max}$), and the density $f$ in the rigth hand side of the previous equation being defined in~\eqref{eq:ind-hyp}.
Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:hat_pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability 
\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i, x_i].\]

@@

We detail in this section the metric considered to evaluate risk prediction performances. Let us recall that in practice, one considers the case $K=2$, see Section~\ref{sec:K=2}. 
In this context, the predictive marker for the lights model is
\[\hat \cR_i^\text{lights} = \dfrac{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i) \big) \hat f(t^{max}_i, y_i | G_i=0 ; \hat \theta)}.\]
Then, let us denote by $\hat \cR$ the marker under study in the general case, and assume that it is measured once at $t = t^{max}$.
A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
\begin{equation*}
  \cC =\P[\hat \cR_i > \hat \cR_j | T^\star_i < T^\star_j],
\end{equation*}
with $i \neq j$ two independent subjects (which does not depend on $i, j$ under the i.i.d. sample hypothesis). 
In our case,  $T^\star$ is subject to right censoring, so one would typically consider the modified $\cC_\tau$ defined by
\begin{equation*}
  \cC_\tau =\P[\hat \cR_i > \hat \cR_j | T_i < T_j , T_i < \tau],
\end{equation*}
with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}. A Kaplan-Meier estimator for the censoring distribution leads to a nonparametric and consistent estimator of $\cC_\tau$ \citep{uno2011c}, already implemented in the \texttt{Python} package \texttt{lifelines}.
Hence in the following, we consider the C-index metric to assess performances.

@@

The first model we consider as a baseline is the well known Cox PH model with time-independent features, also known in the joint modeling context as the ``landmark'' model, in which we include basic time-independent features extracted from longitudinal processes and computed at the prediction time [REF].
In the Cox PH model introduced in \citet{Cox1972JRSS}, a parameter vector $\beta$ is estimated by minimizing the partial log-likelihood given by
\begin{equation*}
  \ell_n^{\text{cox}}(\beta) = n^{-1} \sum_{i=1}^n \delta_i \big( x_i^\top \beta - \log \sum_{i' : t_{i'} \geq t_i} \text{exp}(x_{i'}^\top \beta) \big).
\end{equation*}
For each time-dependant feature, we include subject-specific random effects, values of longitudinal processes at baseline and at time $t^{max}$, slope of longitudinal processes at time $t^{max}$ and area under of longitudinal processes from baseline up to $t^{max}$. The predictive marker for subject $i$ is the regular risk score~\citep{therneau2000cox}, that is $\hat \cR_i^{\text{cox}} = \exp (x_i^\top\hat\beta^{\text{cox}})$, with 
\[\hat\beta^{\text{cox}} \in \argmin_{\beta \in \R^\vartheta} -\ell_n^{\text{cox}}(\beta).\]
We use the \texttt{Python} package \texttt{tick} for the estimation of the previous quantity.
Ties are handled via the Breslow approximation of the partial likelihood \citep{breslow1972contribution}.

@@

We then consider a penalized version of the Cox model, including now much more time-independent features automatically extracted from longitudinal processes using the \texttt{Python} package \texttt{tsfresh}~\citep{christ2018time}.
The predictive marker for the a subject $i$ is $\hat \cR_i^{\text{cox EN}} = \exp (x_i^\top\hat\beta^{\text{cox EN}})$, with 
\[\hat\beta^{\text{cox EN}} \in \argmin_{\beta \in \R^\vartheta} -\ell_n^{\text{cox}}(\beta) + \xi \big( (1-\eta)\norm{\beta}_1 + \frac{\eta}{2} \norm{\beta}_2^2 \big)\] 
the estimate obtained using elastic net penalization (still with the \texttt{tick} package), and
where $\xi$ is chosen by the a 10-fold cross-validation procedure, for a given $\eta \in [0, 1]$.

@@

A classical extension of the Cox model supposes that features depend on time~\citep{sueyoshi1992semiparametric} (and then loosing the risk proportionality property), that is with our notations
\[\lambda_i(t) = \lambda_0(t) \exp\big(x_i^\top \gamma_0 + y_i(t_i^{max})^\top \beta \big). \]
We use the \texttt{survival} \texttt{R} package~\citep{zhang2018time}, and we consider the natural predictive marker for a subject $i$, that is the corresponding risk score given by
\[\hat \cR_i^\text{cox t-d} = \exp \big(x_i^\top \hat\gamma_0 + y_i(t)^\top \hat\beta \big).\]

@@

We consider a multivariate version of JLCMs implemented in the \texttt{R} package \texttt{lcmm}~\citep{2017_lcmm}.
Contrary to the lights model, there is no shared associations between the longitudinal models and the survival model. Given the group membership, each submodel is assumed to be independent. In this context, the predictive marker for subject $i$ is
\[\hat \cR_i^\text{lcmm} = \dfrac{\pi_{\hat \xi}(x_i) f(t^{max}_i, y_i | G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) f(t^{max}_i, y_i | G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i)\big) f(t^{max}_i, y_i | G_i=0 ; \hat \theta)}.\]

@@

In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, \hat \beta^l, b_i^l) \Big\}.\]
